---
title: "MaxDiff Analysis of MSBA Class Preferences"
subtitle: "Counts, MLE, and Bayesian estimation of the MNL model"
author: "Mengyao Ma"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: false
    df-print: kable
execute:
  warning: false
  message: false
  cache: false
---

```{r setup}
#| include: false
# We deliberately keep the toolbox tiny and the code plain.
# Base R does all the heavy lifting (the likelihood, optim, and the
# Metropolis sampler are written by hand). ggplot2 is only used to draw
# the charts, and knitr::kable to print tidy tables.
library(ggplot2)
library(knitr)

set.seed(42)               # reproducibility for the bootstrap + MCMC
theme_set(theme_minimal(base_size = 12))
```


## Introduction

When you want to know how much people *prefer* one thing over another, the
instinct is to hand them a rating scale: "rate each class from 1 to 10." The
problem is that humans are terrible at this. Some people cluster everything
around 7–8, others spread out, and "an 8" means something different to every
respondent. The numbers end up measuring *how a person uses a scale* as much as
*what they actually prefer*.

**MaxDiff** — also called **Best–Worst Scaling** — sidesteps all of that. On
each screen of the survey the respondent sees a small subset of items (here,
four of the ten classes) and is asked only two questions: which one is the
**best**, and which one is the **worst**. That's it. People are genuinely good
at picking extremes and making trade-offs, so the data we get back is far more
reliable than a rating scale, and it forces real differentiation between items
instead of letting everything bunch up in the middle.

In this analysis the ten items are the ten core MSBA classes at UCSD Rady, and
the respondents are the students who took the survey (you and your classmates).
The goal is to estimate students' relative preference for each class, and to do
it three different ways so we can see what each method buys us:

1. A simple **counts analysis** — the percentage of times each class was picked
   best, minus the percentage of times it was picked worst.
2. The **multinomial logit (MNL)** model fit by **maximum likelihood**, where we
   write down the log-likelihood and hand it to `optim()`.
3. The *same* MNL model fit by **Bayesian methods**, using a random-walk
   Metropolis–Hastings sampler we code from scratch.

Here's what to expect before we even start: the three methods should agree on
the broad ranking of classes — if students love a class, all three should say
so. Where they differ is in *how* they quantify the strength of preference and,
especially, in *how they talk about uncertainty*. The counts analysis gives a
quick number, the MLE gives a number plus a classical standard error, and the
Bayesian version gives a whole distribution we can slice however we like.


## The Data

We read in the two files: the design-and-choices file (every row a slot on a
screen) and the labels file (which maps item numbers 1–10 to class names).

```{r read-data}
# read.csv with check.names = FALSE keeps the original headers; we rename
# right after so the rest of the code is clean (no backticks needed).
dat <- read.csv("maxdiff_design_and_choices.csv",
                fileEncoding = "UTF-8-BOM", stringsAsFactors = FALSE)
names(dat) <- c("record_id", "task", "position", "item", "response")

labels <- read.csv("maxdiff_item_labels.csv",
                   fileEncoding = "UTF-8-BOM", stringsAsFactors = FALSE)
labels <- labels[, 1:2]               # the file has some trailing empty columns
names(labels) <- c("item", "label")
labels <- labels[!is.na(labels$item), ]

head(dat, 8)
```

### A wrinkle worth catching

Before we trust the file, it pays to look at how many rows each task has. A
standard MaxDiff screen shows four items, so every task should have four rows.

```{r inspect-tasks}
key <- paste(dat$record_id, dat$task, sep = "_")
rows_per_task <- table(table(key))     # distribution of #rows per task
rows_per_task
```

There are two kinds of tasks in here. Most tasks have four rows — those are the
classic MaxDiff screens. But a chunk of tasks have only **two** rows, and if you
look at them you'll find they always pair one real class against a mysterious
**item 11** that doesn't even appear in the labels file:

```{r show-anchor}
anchor_rows <- dat[dat$item == 11, ]
head(anchor_rows, 4)
nrow(anchor_rows)
```

These two-item screens are **anchoring tasks** (a "dual-response" extension of
MaxDiff, where item 11 is a reference/anchor). They answer a different question
— roughly, "is this class above or below some bar?" — and the assignment's MNL
setup (four items, one best, one worst, ten real classes) is about the standard
screens only. So we filter the anchoring tasks out and keep just the four-item
MaxDiff tasks. Everything from here uses that clean subset.

```{r filter-standard}
# Keep only tasks that have exactly 4 rows (the standard MaxDiff screens).
counts_by_key <- table(key)
std_keys <- names(counts_by_key)[counts_by_key == 4]
d4 <- dat[key %in% std_keys, ]

# Sanity guard: the standard tasks should never contain item 11.
stopifnot(all(d4$item <= 10))
```

### Descriptives

```{r descriptives}
N        <- length(unique(d4$record_id))           # respondents
n_tasks  <- length(unique(paste(d4$record_id, d4$task)))
T_per    <- n_tasks / N                             # standard tasks per person
J        <- 4                                        # items per screen
n_items  <- length(unique(d4$item))                  # total items in the study
```

After filtering we have **`r N`** respondents, each completing **`r T_per`**
standard MaxDiff tasks, for **`r n_tasks`** tasks in total. Every screen shows
**`r J`** of the **`r n_items`** classes.

**Sanity check.** Each task should have exactly one "best" (`response == 1`) and
exactly one "worst" (`response == -1`), with the other two slots at `0`. Let's
confirm that's true for every single task:

```{r sanity}
sp <- split(d4, paste(d4$record_id, d4$task))

ok_best  <- sapply(sp, function(z) sum(z$response ==  1) == 1)
ok_worst <- sapply(sp, function(z) sum(z$response == -1) == 1)

cat("Tasks with exactly one best :", sum(ok_best),  "out of", length(sp), "\n")
cat("Tasks with exactly one worst:", sum(ok_worst), "out of", length(sp), "\n")
```

Both pass for all `r n_tasks` tasks. Finally, because a good MaxDiff design
balances how often each item is shown, the exposure counts should all be roughly
equal:

```{r exposure}
shown <- table(factor(d4$item, levels = 1:10))
exposure_tbl <- data.frame(
  item        = 1:10,
  label       = labels$label[match(1:10, labels$item)],
  times_shown = as.integer(shown)
)
kable(exposure_tbl, caption = "Times each class was shown across the study.")
```

The counts land in a tight band, exactly what a balanced design should produce —
nothing is over- or under-exposed in a way that would bias the analysis.

```{r short-labels}
# Short labels for the charts (the full course titles are too long for axes).
short <- c(
  "403 AI Math & Prog (Nijs)",
  "464 SQL (Nijs)",
  "451 Mktg/Fin/Ops",
  "452 Large Data (Hansen)",
  "453 Business Analytics (August)",
  "444 Analytics Consulting (Peterson)",
  "455 Customer Analytics (Nijs)",
  "454 Capstone (Advisor)",
  "495 Marketing Analytics (Yavorsky)",
  "457 BI Systems (Schibler)"
)
```


## Counts Analysis

The counts analysis is the back-of-the-envelope version, and it's surprisingly
informative. For each class $j$ we ask two questions: of the times it was shown,
what fraction did it get picked as **best**, and what fraction as **worst**? The
difference is a simple score:

$$
\text{score}_j \;=\; \%\text{best}_j \;-\; \%\text{worst}_j
$$

A score near $+1$ means the class is almost always the favorite on screens where
it appears; near $-1$ means it's almost always the goat; near $0$ means it's
either middle-of-the-pack or polarizing (picked best by some, worst by others).
The nice thing is there's no model here at all — we're just counting.

```{r counts}
best_n  <- table(factor(d4$item[d4$response ==  1], levels = 1:10))
worst_n <- table(factor(d4$item[d4$response == -1], levels = 1:10))

shown_n <- as.numeric(shown)
pct_best  <- as.numeric(best_n)  / shown_n
pct_worst <- as.numeric(worst_n) / shown_n
score     <- pct_best - pct_worst

counts_df <- data.frame(
  item       = 1:10,
  label      = labels$label[match(1:10, labels$item)],
  short      = short,
  shown      = shown_n,
  best       = as.integer(best_n),
  worst      = as.integer(worst_n),
  pct_best   = round(pct_best, 3),
  pct_worst  = round(pct_worst, 3),
  score      = round(score, 3)
)
counts_df <- counts_df[order(-counts_df$score), ]
kable(counts_df[, c("item","label","shown","best","worst",
                     "pct_best","pct_worst","score")],
      row.names = FALSE,
      caption = "Counts analysis, ranked from most to least preferred.")
```

### Bootstrap confidence intervals

A single score is a point estimate; we'd like to know how *stable* it is. We get
that with a **bootstrap over respondents**: resample the `r N` students with
replacement, recompute every score, repeat a thousand times, and read off the
2.5th and 97.5th percentiles. Resampling *respondents* (not individual rows)
respects the fact that one student's answers are correlated with each other.

```{r bootstrap-counts}
B <- 1000
resp_ids <- unique(d4$record_id)
id_rows  <- split(seq_len(nrow(d4)), d4$record_id)   # row indices per person

boot <- matrix(NA_real_, B, 10)
for (b in 1:B) {
  samp <- sample(resp_ids, replace = TRUE)
  idx  <- unlist(id_rows[samp])
  db   <- d4[idx, ]
  sh <- as.numeric(table(factor(db$item,                      levels = 1:10)))
  bb <- as.numeric(table(factor(db$item[db$response ==  1],   levels = 1:10)))
  ww <- as.numeric(table(factor(db$item[db$response == -1],   levels = 1:10)))
  boot[b, ] <- bb / sh - ww / sh
}
ci_lo <- apply(boot, 2, quantile, 0.025)
ci_hi <- apply(boot, 2, quantile, 0.975)
```

```{r counts-plot}
#| fig-height: 5
plot_df <- data.frame(short = short, score = score,
                      lo = ci_lo, hi = ci_hi)
plot_df$short <- factor(plot_df$short, levels = plot_df$short[order(plot_df$score)])

ggplot(plot_df, aes(x = score, y = short)) +
  geom_col(fill = "steelblue") +
  geom_errorbarh(aes(xmin = lo, xmax = hi), height = 0.3, color = "grey30") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  labs(x = "Counts score  (% best − % worst)", y = NULL,
       title = "Counts analysis with 95% bootstrap CIs")
```

**What we see.** One class stands head and shoulders above the rest:
**MGTA 495 Marketing Analytics (Yavorsky)**, with a score around $+0.5$ — it gets
picked best far more often than worst, and its confidence interval is nowhere
near anyone else's. Below it sits a cluster of well-liked classes
(453 Business Analytics, 455 Customer Analytics, 452 Large Data) whose intervals
overlap, so their *relative* order is fuzzy. At the bottom,
**444 Analytics Consulting** is the clear least-favorite. A mild surprise is that
**403 AI Math & Programming** — the foundational summer course — lands near the
bottom; that's plausibly less about the material being disliked and more about
intensive prerequisite courses simply being less *fun* to sit through. Take the
instructor-level readings with a grain of salt; this is a small, self-selected
sample.


## From MaxDiff Data to MNL Choices

The counts analysis throws away information. When a class is *shown* but picked
neither best nor worst, that's a signal — it lost to something and beat
something — and counts ignores it. The **multinomial logit (MNL)** model uses
all of it, by modeling the choice itself.

The idea: give each class $j$ a **utility** $\beta_j$. On a screen showing items
$\{a,b,c,d\}$, the probability that $b$ gets picked as **best** is the soft-max
(higher utility ⇒ more likely to be chosen):

$$
P(\text{best} = b) \;=\; \frac{\exp(\beta_b)}{\exp(\beta_a)+\exp(\beta_b)+\exp(\beta_c)+\exp(\beta_d)}
$$

Now here's the clever part of Best–Worst scaling. After picking the best, the
respondent picks the **worst** from the *remaining three*. Picking the worst is
just picking the best on **flipped** utilities — the item you like least is the
one with the lowest $\beta$, i.e. the highest $-\beta$:

$$
P(\text{worst} = c \mid \text{best} = b) \;=\; \frac{\exp(-\beta_c)}{\exp(-\beta_a)+\exp(-\beta_c)+\exp(-\beta_d)}
$$

So **every task gives us two MNL observations**: a best-pick from all four, and a
worst-pick from the remaining three. With `r n_tasks` tasks that's
`r 2*n_tasks` choice observations feeding the likelihood.

**Identifiability.** The soft-max doesn't change if you add the same constant to
every $\beta$ (the constant cancels top and bottom). So only $J-1 = 9$ of the ten
utilities are pinned down. We fix a reference: set $\beta_1 = 0$ (item 1,
403 AI Math) and estimate $\beta_2,\dots,\beta_{10}$ relative to it. Everything is
read as "utility *relative to* 403 AI Math."

We reshape the data into the form the likelihood wants — for each task, the four
items shown, which was best, and which was worst — and precompute a few matrices
so the likelihood runs fast:

```{r reshape}
sp <- split(d4, paste(d4$record_id, d4$task))

# n_tasks x 4 matrix of the item numbers shown on each task
IT <- t(sapply(sp, function(z) z$item))
# which item was best / worst on each task
best_item  <- sapply(sp, function(z) z$item[z$response ==  1])
worst_item <- sapply(sp, function(z) z$item[z$response == -1])
# the 3 remaining items per task (full set minus the best pick)
REM <- t(sapply(seq_along(sp), function(t) setdiff(IT[t, ], best_item[t])))

dim(IT); dim(REM)
```


## MNL via Maximum Likelihood

Summing the log-probabilities of both picks over every task gives the
log-likelihood:

$$
\ell(\boldsymbol{\beta}) \;=\; \sum_{\text{tasks}}
\Big[\, \beta_{j^*_b} - \log\!\!\sum_{j \in \text{shown}}\!\! \exp(\beta_j)
\;-\; \beta_{j^*_w} - \log\!\!\!\!\sum_{j \in \text{shown}\setminus\{j^*_b\}}\!\!\!\! \exp(-\beta_j) \,\Big]
$$

where $j^*_b$ and $j^*_w$ are the best and worst picks on that task. The two
"$-\log\sum\exp$" terms are the soft-max denominators. We compute them with the
**log-sum-exp** trick (`max + log(sum(exp(x - max)))`) so nothing overflows.

```{r loglik}
# numerically-stable log-sum-exp, applied to each row of a matrix
row_lse <- function(M) {
  m <- apply(M, 1, max)
  m + log(rowSums(exp(M - m)))
}

# par is length 9 (betas for items 2..10); item 1 is pinned to 0.
loglik <- function(par) {
  beta <- c(0, par)                       # length 10, reference = item 1

  U  <- matrix(beta[IT],  ncol = 4)       # utilities of the 4 shown items
  ll_best  <- beta[best_item]  - row_lse(U)

  UW <- matrix(-beta[REM], ncol = 3)      # FLIPPED utilities of remaining 3
  ll_worst <- (-beta[worst_item]) - row_lse(UW)

  sum(ll_best + ll_worst)
}
```

We maximize it. `optim` minimizes by default, so `fnscale = -1` tells it to
maximize instead, and `hessian = TRUE` gives us the curvature we need for
standard errors.

```{r optim}
fit <- optim(par = rep(0, 9), fn = loglik,
             method = "BFGS", hessian = TRUE,
             control = list(fnscale = -1))

beta_hat <- c(0, fit$par)                         # full length-10 vector
# SEs from the inverse of the negative Hessian (observed information)
se_free  <- sqrt(diag(solve(-fit$hessian)))
se_hat   <- c(0, se_free)                          # reference item has SE 0
```

The raw $\beta$'s are on a log-odds scale and hard to read. We convert them into
**shares of preference**, which answer a concrete question — *"if all ten classes
appeared on one screen, what fraction of the picks would each one get?"*:

$$
\hat s_j \;=\; \frac{\exp(\hat\beta_j)}{\sum_{k=1}^{10}\exp(\hat\beta_k)}
$$

```{r mle-table}
share_mle <- exp(beta_hat) / sum(exp(beta_hat))

mle_df <- data.frame(
  item   = 1:10,
  label  = labels$label[match(1:10, labels$item)],
  beta   = round(beta_hat, 3),
  se     = round(se_hat, 3),
  share  = round(share_mle, 3)
)
mle_df <- mle_df[order(-mle_df$beta), ]
kable(mle_df, row.names = FALSE,
      caption = "MLE estimates. Item 1 (403 AI Math) is the reference, beta = 0.")
```

```{r mle-plot}
#| fig-height: 5
mp <- data.frame(short = short, share = share_mle)
mp$short <- factor(mp$short, levels = mp$short[order(mp$share)])
ggplot(mp, aes(x = share, y = short)) +
  geom_col(fill = "darkorange") +
  labs(x = "MLE share of preference", y = NULL,
       title = "MNL shares of preference (maximum likelihood)")
```

**MLE vs. counts.** The two rankings come out **identical** here — same order
from top to bottom. That's reassuring but not guaranteed: counts and MNL can
disagree, usually in the muddled middle where utilities are close together and a
few flipped picks reorder things. What the MNL adds is *interpretability and
rigor*: the shares put everything on a single 0–1 scale that sums to 100%, and
every shown-but-not-picked item still contributes information to the estimate.
The standout, **495 Marketing Analytics**, grabs roughly a quarter of all
preference share on its own — more than double the next class.


## MNL via Bayesian Estimation

The MLE gives us a point estimate and a standard error. The Bayesian version
gives us the entire **posterior distribution** of $\boldsymbol{\beta}$, which makes
uncertainty on *derived* quantities (like shares) almost free. By Bayes' rule the
posterior is proportional to likelihood times prior:

$$
\pi(\boldsymbol{\beta}\mid \boldsymbol{y}) \;\propto\; L(\boldsymbol{\beta})\cdot \pi(\boldsymbol{\beta})
$$

We use a weakly informative prior, $\boldsymbol{\beta}\sim N(\mathbf{0},\,10\cdot I)$
— a standard deviation of about $3.2$ on each coefficient, which is plenty diffuse
(it barely nudges the data) without being absurd.

```{r logpost}
log_post <- function(par) {
  loglik(par) + sum(dnorm(par, mean = 0, sd = sqrt(10), log = TRUE))
}
```

### The sampler

A **random-walk Metropolis** sampler is about as simple as MCMC gets. From the
current $\boldsymbol{\beta}$, propose a small random jump; if the proposal has
higher posterior density, always accept it; if lower, accept it with probability
equal to the density ratio. Over many iterations the chain spends time in each
region of parameter space in proportion to its posterior probability.

```{r mh-sampler}
run_mh <- function(start, n_iter, step) {
  K      <- length(start)
  draws  <- matrix(NA_real_, n_iter, K)
  cur    <- start
  lp     <- log_post(cur)
  accept <- 0
  for (t in 1:n_iter) {
    prop  <- cur + rnorm(K, 0, step)        # random-walk proposal
    lp_p  <- log_post(prop)
    if (log(runif(1)) < lp_p - lp) {        # accept/reject on the log scale
      cur <- prop; lp <- lp_p; accept <- accept + 1
    }
    draws[t, ] <- cur
  }
  list(draws = draws, accept_rate = accept / n_iter)
}
```

**Tuning the step size.** If the jumps are tiny the chain accepts almost
everything but crawls; if they're huge it almost never accepts and gets stuck.
The sweet spot is an acceptance rate of roughly 25–40%. A step of `0.06` lands us
right in that band here. (Rule of thumb: too low ⇒ halve the step; too high ⇒
double it.)

```{r run-mh}
n_iter <- 20000
mh <- run_mh(start = rep(0, 9), n_iter = n_iter, step = 0.06)
cat("Acceptance rate:", round(mh$accept_rate, 3), "\n")

burn <- floor(0.25 * n_iter)                 # discard first 25% as burn-in
post <- mh$draws[(burn + 1):n_iter, ]
post_full <- cbind(0, post)                  # add the reference column (beta_1 = 0)
```

### Diagnostics

A healthy chain looks like a **fuzzy caterpillar** — it bounces around a stable
level rather than drifting. Here are trace plots for three of the parameters:

```{r trace}
#| fig-height: 4.5
idx_show <- c(8, 4, 5)   # free-param columns -> items 9, 5, 6
labs_show <- c("item 9 (495 Mktg)", "item 5 (453 Bus Analytics)",
               "item 6 (444 Consulting)")
trace_df <- data.frame(
  iter  = rep(1:n_iter, 3),
  value = c(mh$draws[, idx_show[1]], mh$draws[, idx_show[2]], mh$draws[, idx_show[3]]),
  param = factor(rep(labs_show, each = n_iter), levels = labs_show)
)
ggplot(trace_df, aes(iter, value)) +
  geom_line(linewidth = 0.2, color = "grey30") +
  geom_vline(xintercept = burn, linetype = "dashed", color = "red") +
  facet_wrap(~param, ncol = 1, scales = "free_y") +
  labs(x = "iteration", y = expression(beta),
       title = "Trace plots (red line = end of burn-in)")
```

All three settle into a stable band almost immediately and then just wiggle —
no slow drift, no getting stuck. The chain has converged.

### Posterior summaries

```{r bayes-summaries}
post_mean  <- colMeans(post_full)
post_ci    <- apply(post_full, 2, quantile, c(0.025, 0.975))

# Push EVERY draw through the soft-max to get a posterior over shares.
share_draws <- t(apply(post_full, 1, function(b) exp(b) / sum(exp(b))))
share_mean  <- colMeans(share_draws)
share_ci    <- apply(share_draws, 2, quantile, c(0.025, 0.975))

bayes_df <- data.frame(
  item       = 1:10,
  label      = labels$label[match(1:10, labels$item)],
  post_mean  = round(post_mean, 3),
  ci_lo      = round(post_ci[1, ], 3),
  ci_hi      = round(post_ci[2, ], 3),
  share_mean = round(share_mean, 3)
)
bayes_df <- bayes_df[order(-bayes_df$post_mean), ]
kable(bayes_df, row.names = FALSE,
      caption = "Posterior means, 95% credible intervals, and mean shares.")
```

```{r bayes-plot}
#| fig-height: 5
bp <- data.frame(short = short, share = share_mean,
                 lo = share_ci[1, ], hi = share_ci[2, ])
bp$short <- factor(bp$short, levels = bp$short[order(bp$share)])
ggplot(bp, aes(x = share, y = short)) +
  geom_col(fill = "seagreen") +
  geom_errorbarh(aes(xmin = lo, xmax = hi), height = 0.3, color = "grey30") +
  labs(x = "Posterior mean share of preference", y = NULL,
       title = "MNL shares of preference (Bayesian) with 95% credible intervals")
```

**Bayes vs. MLE.** The posterior means land almost exactly on top of the MLE
point estimates, and the posterior standard deviations match the MLE standard
errors to two decimals. That's exactly what theory predicts: with a fair amount
of data and a diffuse prior, the likelihood dominates and the two approaches
converge. The difference isn't in the *numbers* — it's in what we can do with
them. We got 95% intervals on the **shares** for free, just by pushing each draw
through the soft-max and taking quantiles. Getting that from the MLE would
require the delta method and some calculus.


## Comparing the Three Methods

```{r compare-table}
ord_counts <- order(-score)
ord_mle    <- order(-share_mle)
ord_bayes  <- order(-share_mean)

rank_counts <- match(1:10, ord_counts)
rank_mle    <- match(1:10, ord_mle)
rank_bayes  <- match(1:10, ord_bayes)

compare <- data.frame(
  label        = labels$label[match(1:10, labels$item)],
  counts_score = round(score, 3),
  counts_rank  = rank_counts,
  mle_share    = round(share_mle, 3),
  mle_rank     = rank_mle,
  bayes_share  = round(share_mean, 3),
  bayes_rank   = rank_bayes
)
compare <- compare[order(compare$mle_rank), ]
kable(compare, row.names = FALSE,
      caption = "Side-by-side comparison, sorted by MLE rank.")
```

A visual version makes the agreement obvious. To let you trace a single class
across all three panels, every panel is sorted by the **MLE rank**, and each
class keeps its own color:

```{r compare-plot}
#| fig-height: 6
#| fig-width: 9
# Order classes by MLE rank; reuse that ordering in all three panels.
ord <- ord_mle
lab_ord <- factor(short[ord], levels = rev(short[ord]))

long <- rbind(
  data.frame(short = short[ord], value = score[ord],      method = "Counts (score)"),
  data.frame(short = short[ord], value = share_mle[ord],  method = "MLE (share)"),
  data.frame(short = short[ord], value = share_mean[ord], method = "Bayes (share)")
)
long$short  <- factor(long$short, levels = rev(short[ord]))
long$method <- factor(long$method,
                      levels = c("Counts (score)", "MLE (share)", "Bayes (share)"))

ggplot(long, aes(x = value, y = short, fill = short)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~method, scales = "free_x") +
  labs(x = NULL, y = NULL,
       title = "Three methods, one ranking — classes sorted by MLE rank") +
  theme(axis.text.y = element_text(size = 8))
```

**Where they agree.** Everywhere, basically. All three methods put
**495 Marketing Analytics** firmly on top and **444 Analytics Consulting** firmly
at the bottom, and in this dataset the full ranking happens to be identical
across methods. Agreement at the extremes is almost guaranteed — a class that's
genuinely loved or hated shows up that way no matter how you slice the data.

**Where they could disagree.** The risk is always in the middle, where several
classes have nearly equal utility (look how the green credible intervals overlap
for the middle cluster). There, a handful of flipped picks can reorder ranks, and
the three methods might land in different orders. We got lucky with a clean
ranking, but those middle ranks are the "soft" ones — don't over-read a one-spot
difference there.

**What the MNL adds over counts.** It models the choice process explicitly. Every
*shown* item — even one that wasn't picked — contributes information about *not*
being chosen, which the counts score throws away. The result is estimates on a
principled scale (shares that sum to 100%) with proper standard errors.

**What Bayes adds over MLE.** Uncertainty on *derived* quantities, for free. Want
a credible interval on a share, or on "class A's share minus class B's," or on a
market-simulation scenario? Just compute that quantity for every MCMC draw and
take quantiles. From the MLE you'd need the delta method for each new quantity.


## Discussion

**The substantive finding.** If you're a student deciding what to look forward
to, the data is loud about one thing: **MGTA 495 Marketing Analytics with
Yavorsky** is the standout favorite — it pulls roughly a quarter of all
preference share on its own. Behind it sits a solid tier of well-liked electives
(Business Analytics, Customer Analytics, Large Data). The least-loved class is
**MGTA 444 Analytics Consulting**, and the foundational summer course
**403 AI Math & Programming** also lands low — which probably says more about
intensive prerequisite courses being less *enjoyable* than about the material
being bad. Two honest caveats apply throughout: this is a small, **self-selected**
sample of MSBA students, and things like presentation order, prior coursework,
and where someone is in the program can all color responses. So read these as
*this cohort's vibe*, not a universal verdict.

**The methodological finding.** Each method has its moment:

- **Counts** — the quick descriptive read. Perfect for an executive dashboard or
  a first look; no modeling assumptions, instantly explainable.
- **MLE MNL** — rigorous point estimates with classical standard errors. Reach
  for it when you need a single defensible number to report.
- **Bayesian MNL** — when you care about uncertainty on *derived* quantities
  (shares, "what if we added a new class?" market simulations, differences
  between classes), or when you want to extend the model **hierarchically** to
  estimate each student's individual preferences rather than just the group
  average.

**What I'd explore next.** The obvious extension is exactly that hierarchical
model: a per-respondent set of $\beta$'s drawn from a population distribution,
which would let us see *who* loves which class instead of only the average — and
would naturally fold the anchoring tasks (item 11) we set aside back into the
analysis. I'd also like to bring those anchor screens in properly, since they
carry information about whether a class clears an absolute bar, not just how it
stacks up against three neighbors.
